\chapter{Multilevel K-Way Graph Partitioning}
This chapter presents the Multilevek K-way Graph Partitioning algorithm which is one of the most widespread GP algorithms. Throughout each section METIS own implementation will be mentioned both for the edge cut and communication volume metric. 

\section{Why Multilevel K-way?}
There are many algorithms that are able to produce good partitions.The techniques vary from global methods that partition the graph directly, to breadth first search to methods from computational geometry and linear algebra~\cite{buluc_recent_2016}. The main idea behind multilevel k-way partitioning is to coarsen down the graph to a smaller graph first. On this smaller graph an initial partition is performed on the small graph before it is uncoarsened back up to its original size. For each uncoarsening level the partition is refined. The next sections in this chapter will go in detail on how METIS implements this approach.




%Explain how multilevel works briefly, point to next sections that will go in detail
%Examples in how it outperforms other algorithms. O(E) time complexity for example, drastically faster than the rest
%(maybe specify kway its better, as its faster than bisection. By a factor of O(logk))
%Metis/ParMetis and Scotch are widely used implementations of multilevel kway 
%Maybe a sentence what others sacrifice (spectral methods are expensive, geometric methods need coordinates and often compromise quality) makes the multilevel k-way choice feel motivated rather than just asserted.

Most tools and algorithms implement the \textit{multilevel} heuristic~\cite{GilbertMichaelS.2024JMGP}. Its popularity can be attributed to the fact that this approach tends to perform better than others for large graphs~\cite{buluc_recent_2016, KarypisGeorge1999PMkP, KarypisGeorge1998AFaH}.






This section will present the multilevel approach as the most important in parallel computing\\
Coarsening phase\\
Initial partition \\
Uncoarsening and refinement\\
All the way with examples of how Metis is implemented, especially for the volume objective. 