\chapter{Greedy Refinement Using Communication Volume} \label{chapter5}
This chapter discusses how Greedy Refinement (GR) works using the total communication volume objective function (VOL) and our partition-aware communication volume objective function NVOL. While the GR algorithm itself is well-established and used by partitioners such as METIS, its integration with NVOL is a contribution of this work. Building on the gain calculations from Chapter~\ref{cv_calculation}, we show how GR uses these gains to decide which vertex moves to perform.

The first section presents the logic used for GR with VOL, the standard communication volume objective function used by partitioners such as METIS. The second section describes how GR works with NVOL. Each section concludes with an algorithm summarizing how GR evaluates vertex moves under the corresponding objective.

\section{GR with VOL}\label{sec:METIS refinement}
In this section, we describe how GR works when using VOL as the objective function. The goal of VOL is to minimize the total communication volume of the graph. After the preparation phase is completed, refinement starts: using GR, the partitioner visits all the vertices in the boundary set.  

For each boundary vertex $i$, all the neighboring partitions $N_i$ are visited, checking the corresponding value in the $G_i$ vector. A vertex move is only considered if removing $i$ from its current partition does not violate the minimum partition weight. Among all considered moves, GR selects the one with the highest gain in $G_i$ that is allowed by the balance constraints, i.e., the move that reduces the total communication volume the most. To prioritize high-gain moves, METIS relaxes the balance constraint in proportion to the gain. Using a constant called the fudge factor $f$, a move to partition $p$ is allowed to exceed the maximum partition weight by $f\cdot G_i[p]$. This allows moves that improve the communication volume significantly, at the cost of a few extra vertices in a partition.

When two moves have the same gain, METIS breaks ties using two further criteria: external degree (see Section~\ref{sec:preparation for refinement}) and partition balance. It first chooses the move with the highest specific external degree. Since all vertices within a coarse vertex are assigned to the same partition upon projection (see Section~\ref{sec:preparation for refinement}), a move toward a partition with high external degree may reduce volume at finer levels. If instead two candidates have the same gain and the same external degree, the move to the more underweight partition is chosen. This improves the weight balance among the partitions. Algorithm~\ref{alg:ref_vol} shows the logic GR with VOL applies to each boundary vertex $i$. 

\begin{algorithm}[H]
	\caption{GR with VOL}\label{alg:ref_vol}
	\begin{algorithmic}
		\State \textbf{Input:} boundary vertex $i$ under consideration, with neighbor partitions $N_i[0 \ldots n{-}1]$
		\State \Comment{$N_i[k]$ denotes the $k$-th neighbor partition of $i$}
		\If{$w(P[i]) - w(i) \geq W^{\min}$}
		\For{$k = n{-}1$ \textbf{downto} $0$}  \Comment{Find first feasible candidate}
		\If{$G_i[N_i[k]] \geq 0$ \textbf{AND}\\	
			\hspace*{4.1em}$w(N_i[k]) + w(i) \leq W^{\max} + f \cdot G_i[N_i[k]]$}
		\State \textbf{break}
		\EndIf
		\EndFor
		\If{$k < 0$} \textbf{continue} \Comment{Go to next vertex, no feasible candidate} \EndIf
		
		\For{$j = k{-}1$ \textbf{downto} $0$}  \Comment{Search for a better candidate}
		\State $p_j \gets N_i[j],\; p_k \gets N_i[k]$
		\If{$G_i[p_j] > G_i[p_k]$ \textbf{AND} \\
			\hspace*{4.1em}$w(p_j) + w(i) \leq W^{\max} + f \cdot G_i[p_j]$}
		\State $k \gets j$ \Comment{Better gain, relaxed balance}
		\ElsIf{$G_i[p_j] = G_i[p_k]$ \textbf{AND}\\
			\hspace*{6.3em}$ED_{p_j}[i] > ED_{p_k}[i]$ \textbf{AND}\\
			\hspace*{6.3em}$w(p_j) + w(i) \leq W^{\max}$}
		\State $k \gets j$ \Comment{Same gain, better connectivity}
		\ElsIf{$G_i[p_j] = G_i[p_k]$ \textbf{AND} \\
			\hspace*{6.3em}$ED_{p_j}[i] = ED_{p_k}[i]$ \textbf{AND} \\
			\hspace*{6.3em}$w(p_j) < w(p_k)$}
		\State $k \gets j$ \Comment{All else equal, prefer underloaded partition}
		\EndIf
		\EndFor
		\State Move $i$ to $N_i[k]$
		\EndIf
	\end{algorithmic}
\end{algorithm}

\section{GR with NVOL}\label{sec:NVOL refinement}
While GR itself follows the same overall patter as the VOL variant described above, its adaptation to NVOL is a contribution of this work. The goal of NVOL is to minimize the maximum outgoing communication volume ($MV^{\mathrm{max}}$) held by any single partition. In Section~\ref{Partition specific gain calculation}, we explained how the gain table $SG_i$ is computed for a boundary vertex $i$. To briefly recall, each entry $SG_i[p][j]$ records the change in outgoing communication volume that partition $p$ would experience if vertex $i$ were moved to the neighboring partition $j$. The rows of the table correspond to affected partitions, while the columns correspond to possible destinations. 

The GR algorithm follows the same pattern as before. For each boundary vertex $i$, all the neighboring partitions $N_i$ are visited, checking the corresponding values of the $SG_i$ table. Here too, moves are only considered if removing $i$ from its current partition does not violate the minimum partition weight. As with VOL, NVOL also enforces partition max weight constraints: any move must keep the destination partition's weight within the allowed bounds. For a potential move to a destination partition $j$, NVOL computes the highest volume that any affected partition would hold after the move. If any partition exceeds the current global $MV^{\mathrm{max}}$, the move is immediately discarded. 

NVOL will look first for a candidate move that does not increase the global $MV^{\mathrm{max}}$. As is further discussed in Section~\ref{sec:encourage more moves}, this might lead to an increase in local $MV^{\mathrm{max}}$. After the first candidate has been found, NVOL looks for a better option ranked according to three criteria, applied in the following order of priority: local $MV^{\mathrm{max}}$, external degree, and balance. Under the first criterion, NVOL seeks a move that minimizes the local top volume. Among the possible destinations, NVOL chooses the destination $j$ that results in the lowest top volume across the remaining affected partitions, provided that this is lower than the previous candidate. When two moves have the same local top volume, NVOL favors destinations toward which vertex $i$ has a high external degree as a tiebreaker. This tiebreaker follows the same logic as in VOL: it favors moves where the vertex is strongly connected to the destination partition. The final criterion is used when the previous two are tied. NVOL prefers moving vertices to lighter partitions, improving the weight distribution among the partitions. 

Just like VOL, NVOL uses the fudge factor to influence the weight constraint; however, it is used slightly differently. Section~\ref{sec:encourage more moves} explains how the fudge factor is exploited by NVOL to promote and prohibit moves at the same time. 

Algorithm~\ref{alg:ref_nvol} shows how GR chooses a vertex move using NVOL. The algorithm consists of a main loop iterating over all possible destinations of the boundary vertex. In an inner loop, it computes the highest volume after the move. To do this, it scans all entries $SG_i[k][j]$ for $k \in N_i \cup \{P[i]\}$ and computes $PV[k] + SG_i[k][j]$, where $PV[k]$ is the volume partition $k$ holds before the move. We include $P[i]$ in the loop to also record how the source partition is affected by the move. The partition achieving the highest post-move volume is recorded as the local top partition, with volume \texttt{tmpVol} and partition id \texttt{tmpPid}. The remaining if-else branches encode the three ranking criteria described above. At the end, if a candidate has been found, the move is performed.

\begin{algorithm}[h]
	\caption{GR with NVOL}\label{alg:ref_nvol}
	\begin{algorithmic}
		\State \textbf{Input:} boundary vertex $i$ under consideration, with neighbor partitions $N_i[0 \ldots n{-}1]$, global maximum partition volume $MV^{\mathrm{max}}$
		\If{$w(P[i]) - w(i) \geq W^{\min}$}
			\State candidateVol $\gets -1 $
			\State destination $\gets -1 $
			\State Compute gain table $SG_i$
			\ForAll{$j \in N_i$}
				\State tmpVol $\gets 0$
				\State tmpPid $\gets -1$
				\ForAll{$k \in N_i \cup P[i]$}
					\If{tmpVol$< PV[k] + SG_i[k][j]$}
						\State tmpVol $\gets PV[k] + SG_i[k][j]$
						\State tmpPid $\gets k$
					\EndIf
				\EndFor
				\If{candidateVol = -1 \textbf{AND}\\
				\hspace*{4.1em}$MV^{\mathrm{max}} \geq$ tmpVol \textbf{AND} \\
				\hspace*{4.1em}$w(j) + w(i) \leq W^{\max} + f \cdot SG_i[\mathrm{tmpPid}][j]$}
					\State destination $\gets j$ 						\Comment{Move does not increase $MV^{\mathrm{max}}$}
					\State candidateVol $\gets$ tmpVol
			
				\ElsIf{candidateVol $\neq -1$ \textbf{AND}\\
				\hspace*{6.3em}candidateVol > tmpVol \textbf{AND}\\
				\hspace*{6.3em}$w(j) + w(i) \leq W^{\max} + f \cdot SG_i[\mathrm{tmpPid}][j]$}
					\State destination $\gets j$ 						\Comment{Lower local max partition volume after move}
					\State candidateVol $\gets$ tmpVol
			
				\ElsIf{candidateVol $\neq -1$ \textbf{AND}\\
				\hspace*{6.3em}candidateVol = tmpVol \textbf{AND}\\
				\hspace*{6.3em}$ED_{j}[i] > ED_{\mathrm{destination}}[i]$ \textbf{AND}\\
				\hspace*{6.3em}$w(j) + w(i) \leq W^{\max} + f \cdot SG_i[\mathrm{tmpPid}][j]$}
					\State destination $\gets j$ 						\Comment{Same gain, better connectivity}
					\State candidateVol $\gets$ tmpVol
			
				\ElsIf{candidateVol $\neq -1$ \textbf{AND}\\
				\hspace*{6.3em} candidateVol = tmpVol \textbf{AND} \\
				\hspace*{6.3em}$ED_{j}[i] = ED_{\mathrm{destination}}[i]$ \textbf{AND} \\
				\hspace*{6.3em}$w(j) < w(\mathrm{destination})$}
					\State destination $\gets j$						\Comment{All else equal, prefer underloaded partition}
					\State candidateVol $\gets$ tmpVol
				\EndIf
			\EndFor
			\If{destination $\neq -1$}									\Comment{Make sure a candidate has been found}
				\State Move $i$ to destination
			\EndIf
	\EndIf
	\end{algorithmic}
\end{algorithm}

\subsection{Encouraging more moves}\label{sec:encourage more moves}
A deliberate design choice in NVOL was to promote a higher number of vertex moves. Especially when choosing the first candidate, NVOL does not require an immediate improvement of the local top volume. Instead, it chooses any move that leads to a local top below the global maximum. This means that in practice NVOL also chooses moves that increase the local top volume as long as it remains below the global maximum.

The goal of this design is to increase the number of moves NVOL performs during a refinement stage. During testing, we experimented with a stricter criterion that required every accepted move to improve the local top volume. Under this stricter rule, significantly fewer vertices were moved during refinement. On average, the stricter rule performed half as many moves as the more relaxed approach. By encouraging more moves, we observed two effects. First, NVOL was able to explore more solutions: since more moves were performed, more configurations were evaluated. Second, we increased the chance of avoiding getting stuck in a local maximum. If no move could improve the partition's volume, GR would remain stuck, unable to find an improving move. By allowing a ``worsening'' move, we increase the chance of finding better solutions. Since the condition still guarantees that the global maximum is never exceeded, the objective function's value is protected throughout the process.

To prevent this relaxed condition from accepting moves that are too damaging, NVOL incorporates the fudge factor as a safety mechanism. The fudge factor is used to adjust the maximum allowed partition weight; however, in NVOL we multiply it by the gain for the top partition. Specifically, for a vertex $i$, the partition $m$ with the highest local volume, and a destination partition $j$, we multiply the fudge factor by $SG_i[m][j]$. This ensures that moves with a negative impact on the local top partition are more likely to be filtered out by the weight constraint. By making the maximum allowed weight smaller, it becomes harder to accept a move. In the opposite case, it acts as a relaxing factor, allowing moves that improve the local top. In this way, $f$ acts as a soft penalty that discourages harmful moves and promotes positive ones.