\chapter{METIS and Supporting Tooling}\label{chap:NVOL tooling}
This chapter presents the tools developed and used throughout this thesis to produce the results in Chapter~\ref{chap:results}. The first section describes in detail how METIS was adapted to support NVOL, while the second section covers the tools used for data analysis and visualization.

\section{METIS}
For this thesis we used METIS as our multilevel $k$-way partitioner. The METIS framework is open source\footnote{\url{https://github.com/KarypisLab/METIS}}, developed by George Karypis, and widely used. We have also made public a version extended with NVOL\footnote{\url{https://github.com/riccardoie/METIS-with-NVOL}}. Most of the logic behind the multilevel $k$-way algorithm has already been covered in Chapters~\ref{chap:multilevel-kway}-\ref{chapter5}, so we refer the reader to those chapters.

\subsection{Implementing NVOL} \
Implementing NVOL required extending METIS's existing infrastructure. We first defined a new volume-based objective function named NVOL. Because it is volume-based, NVOL reuses the coarsening and initial partitioning code of the total volume objective function (VOL); our extensions are confined to the refinement stage.

%one sentence about how graphs and ctrl are these general global structs
We added two new structs to METIS: one to carry per-partition information, and one for the refinement table. The \texttt{pinfo\_t} struct holds an integer representing a partition's outgoing volume, and METIS's main \texttt{graph\_t} struct now carries a pointer to it. At the start of refinement we allocate an array of $k$ \texttt{pinfo\_t} entries, one per partition, and update them throughout refinement to track partition-specific volumes. The second struct, \texttt{vol\_refinement\_table}, holds a pointer to an integer array called \texttt{gain\_list}; \texttt{graph\_t} instead stores a pointer to a \texttt{vol\_refinement\_table} struct, yielding effectively an array of \texttt{vol\_refinement\_table}. This forms the gain table described in Chapter~\ref{cv_calculation}.

When refinement begins, METIS allocates all of the memory needed by its refinement structures. NVOL reuses these structures and adds the two described above. The per-partition info requires $k$ integers, while the refinement table requires $k \times k$. Allocating a full table for every boundary vertex would be prohibitively expensive, so we instead allocate a single shared refinement table that is repopulated for each vertex under consideration. This keeps NVOL's memory overhead only marginally higher than METIS's.

During the preparation phase at each level, METIS clears its refinement structures and recomputes them. The graph is projected up to a finer level, and all the structs containing information about neighboring vertices, partitions, and edges need to be updated. We added NVOL as a case in all relevant functions. Most of the existing code was reusable, since NVOL shares the same base as VOL; we extended it only to populate \texttt{pinfo\_t}. While evaluating the external degree and the set of neighboring partitions of each vertex, we now also store this information in \texttt{pinfo\_t}. The computational cost of this operation is negligible as the information was already being computed, just not retained. For each vertex visited, we increase its partition's outgoing volume by the vertex size for every unique neighbor partition the vertex has.

The heart of NVOL lives in \texttt{kwayfm.c}, which contains all of the refinement algorithms. The file's dispatching function, \texttt{Greedy\_KWayOptimize}, selects a refinement algorithm based on the objective function; we added a new branch that dispatches to our refinement function, \texttt{Greedy\_KWayNVolOptimize}. This function iterates over the boundary set and decides which vertex moves to perform. The NVOL refinement implementation rests on two functions: \texttt{Greedy\_KWayNVolOptimize} and \texttt{compute\_potential\_move\_volume\_change}. The former is adapted from METIS's existing \texttt{Greedy\_KWayVolOptimize}, retaining the overall priority-queue-driven structure but replacing the move-selection logic to target the NVOL objective function. The latter is a new helper that supplies the per-partition volume estimates required by the modified selection logic.

\subsubsection{Greedy\_KWayNVolOptimize}
\texttt{Greedy\_KWayNVolOptimize} follows the same outer structure as the other greedy refinement algorithms: boundary vertices are inserted into a priority queue ordered by their total communication volume gain. If we wanted to use partition-specific gain here instead we would have had to calculate the refinement table during preparation and then later recalculate it during refinement. To avoid repeating this operation we settled for using the total communication volume gain as a measure, as it was ready to use and it equals the sum of partition-specific gains. Vertices are then extracted one at a time and considered for relocation. NVOL diverges from METIS only in how it chooses the destination partition. This selection is detailed in algorithm~\ref{alg:ref_nvol}.

Each time a vertex is extracted, NVOL populates the gain calculation table by using \texttt{compute\_potential\_move\_volume\_change} (described below). It then iterates over all possible destinations to find the most viable option. Each iteration follows the same pattern: identify the highest local volume that would result from the move, then evaluate the move. Because only the neighbors of the extracted vertex experience a change in volume, NVOL inspects only those vertices via an inner loop. There, \texttt{pinfo\_t} provides the current partition volumes, from which the corresponding gain values in the gain table are subtracted.

After the inner loop, NVOL evaluates the candidate move through a series of conditional checks. If no move has yet been accepted, we simply require that the highest local volume after the move be smaller than the highest recorded volume in the graph. This value is stored in the \texttt{graph\_t} struct, and updated in \texttt{compute\_potential\_move\_volume\_change}. By doing this, we ensure that any accepted move will not increase the graph's maximum per-partition outgoing volume. Once an initial candidate has been accepted, we look for the move that yields the smallest maximum local volume. Ties are broken in two stages: we first prefer destinations towards which the vertex has greater external degree, and if that is also tied we prefer moves toward the more underweight partition. If no candidate is found, NVOL goes on to the next vertex.

Once every destination has been evaluated and a move selected, we update the \texttt{pinfo\_t} information. The remaining machinery (boundary updates, subdomain connectivity tracking, and the rebalancing branch) is left essentially unchanged from the upstream METIS code. This operation will be repeated until the priority queue has been emptied. The greedy refinement is repeated a set number of times, however, if no moves are made or no change in volume occurs, the algorithm exits early; NVOL inherits this behavior.

\subsubsection{compute\_potential\_move\_volume\_change}
\texttt{compute\_potential\_move\_volume\_change} is the helper that makes the selection of moves possible. It implements the algorithm described in Section~\ref{sec:partition specific gain imp}. The function begins by zeroing out the refinement table to ensure no stale values from other vertices remain. Since this pass already touches each of the $k$ rows of the table, we use the opportunity to also record which partition currently has the highest volume; this value defines the ceiling when choosing moves. The double array structure of the refinement table behaves as a matrix: the first index identifies a row and the second a column, so \texttt{vol\_refinement\_table[p].gain\_list[j]} stores the volume change that partition $p$ would experience if a vertex $i$ were moved to partition $j$. To fill this table, the function walks the neighborhood of $i$ and updates the relevant entries of the table according to the first three cases described in Section~\ref{Partition specific gain calculation}. Once all neighbors have been accounted for, the function handles case 4 and the volume contribution of the moving vertex through an additional loop, adjusting the volumes of the source and destination partitions accordingly. 

\subsubsection{Time complexity}
The additional bookkeeping required by NVOL comes at a measurable cost to METIS's runtime. During refinement, every visited vertex triggers a set of additional iterations that are not present in the VOL implementation of METIS. These extra passes accumulate quickly as the graph grows. The result is that NVOL is noticeably slower than VOL, with the gap widening as the input size increases.

The two parameters that drive this slowdown are the average vertex degree and the number of partitions. The added iterations walk over both the neighboring vertices of each visited vertex and the partitions adjacent to it, so denser graphs and finer partitionings each independently inflate the work performed per refinement step. When both quantities are large simultaneously, the effects compound and the runtime penalty becomes substantial.

The source of this overhead becomes clear when we isolate the portion of the code where \texttt{Greedy\_KWayNVolOptimize} and \texttt{Greedy\_KWayVolOptimize} diverge. Central to the NVOL variant is the call to \texttt{compute\_potential\_move\_volume\_change}. This routine has a time complexity of$O(n(n + m))$, where $n$ denotes the number of neighboring partitions of a vertex and $m$ the number of neighboring vertices. Because this work is repeated for every vertex considered during refinement, even a modest constant-factor difference at the per-vertex level translates into a significant aggregate slowdown over a full pass. The same part of code in VOL on the other hand, has a complexity of $O(n)$, simply iterating once over the neighboring partitions of the vertex. 
%
%NVOL's added complexity affects the speed of METIS. For every visited vertex during refinement it adds additional iterations which increase the runtime especially on bigger graphs.  
%
%These new additions iterate over the number of neighboring vertices, and the number of neighboring partitions. This is visible when NVOL is run using a high number of partitions, or generally on a graph with many edges. When both values are high, the slowdown is great. 
%
%If we purely consider the part of \texttt{Greedy\_KWayNVolOptimize} and \texttt{Greedy\_KWayVolOptimize} differ, we confirm why NVOL is slower. Let's start with \texttt{compute\_potential\_move\_volume\_change} which has a time complexity of $O(n(n + m))$. 

%So vertex move chooice is n iterations w n +1 inner loop = O(n²) n is nnbrs

%compute potential has O(n²) to empty the table. Main loop is O(mxn) m is adj vertices. Finishes with O(n).

% O(n(n + m)) + O(n²) 
%
%
%Given the additional internal loop, the move choice code retains a higher time complexity to METIS. NVOL's inner logic has a time complexity of $O(d_i²)$, where $d_i$ represent the number of neighbor partitions a vector $i$ has. This is 

\subsection{Inputs}
METIS requires three inputs: a graph file, the number of partitions, and the objective function. A few optional flags are also available to fine-tune the algorithm.

The graph file must follow a specific structure. The first line is a header whose first two values are the number of vertices and the number of unique edges in the graph. The header may also contain two additional parameters specifying whether the graph has vertex/edge weights and whether it has multiple weights per vertex or edge. Further details can be found in the official manual\footnote{\url{https://github.com/KarypisLab/METIS/blob/master/manual/manual.pdf}}. All graphs used in this work are single-weight, with unit weights on both edges and vertices, a list of them can be found in table~\ref{tab:heart_dataset} and table~\ref{tab:suitesparse}. Note, however, that NVOL also supports graphs with single vertex weights.

The remaining lines of the graph file describe the graph's connectivity: each line corresponds to a vertex, and its entries are the ids of the vertices it is connected to.

\subsection{Output}
Once the algorithm terminates, METIS produces a partition file. This text file contains $n$ lines, where $n$ is the number of vertices in the graph, and each line holds a single integer in the range $[0, k-1]$ indicating the partition the corresponding vertex was assigned to. The partition files are name graphName.part.$k$, where $k$ is the number of partitions. 


\section{Graph Tools} \label{sec:graph tools}
In this section we will present a few simple tools that were developed during this thesis to display and analyze the results produced by the partitioning algorithm as well as graph formatting. The scripts can all be found in this public repository \footnote{\url{https://github.com/riccardoie/Graph-tools/tree/main}}.

\subsection{calc\_partitions}
To verify and analyze the per-partition outgoing and incoming communication volumes produced by our experiments, we wrote a small program, \texttt{calc\_partitions}, that computes them directly from the partition files. The program contains two functions, \texttt{outgoing\_volstats} and \texttt{incoming\_volstats}, each taking two arguments: the original graph filename and the path to the folder containing the partition files. Given these inputs, each function computes its respective volume for every \texttt{.part.$k$} file associated with that graph. The results are written to a text file divided into sections, one per partition file, each introduced by a header of the form \texttt{graphName.part.$k$}. Within each section, the partitions are listed in order of partition id, from $0$ to $k-1$. Each function produces its own output file.

%DEFINITION OF INCOMING HERE MIGHT MOVE TO RESULTS AND INSTEAD ADD (see section in paragraph above)

The outgoing communication volume of a partition was defined in Section~\ref{Partition specific gain calculation} as the sum of the outgoing volumes of the vertices it contains. The incoming volume has not been introduced yet, so we extend the definitions accordingly. If outgoing communication volume measures the data a partition must send out, incoming communication volume measures the data it must receive from outside. Using the notation from Chapter~\ref{cv_calculation}, the incoming communication volume of partition $n$ is
\begin{equation}
	\mathrm{incoming\_volume}[n] = \sum_{\substack{v \in V_b \\ n \in N_v}} w(v),
\end{equation}
where $V_b$ is the set of boundary vertices and $N_v$ is the set of partitions of $v$'s neighbors, excluding $v$'s own partition $P[v]$. In other words, every boundary vertex $v$ contributes its weight $w(v)$ to each partition that are connected to it.

% END OF INCOMING DEF

Both functions share the same traversal. For every partition file, the function initializes an array of length $k$ and then iterates over the graph's vertices, collecting for each vertex the set of distinct partitions among its neighbors that differ from its own. The two functions only diverge in how they update the array from this set: \texttt{outgoing\_volstats} adds the size of the set to the entry for the vertex's own partition, while \texttt{incoming\_volstats} increments by one the entry of every partition in the set. Using a set ensures each external partition is counted at most once per vertex, matching the definition of communication volume from Section~\ref{Partition specific gain calculation}. Once the graph has been fully traversed, the resulting array is stored in a dictionary keyed by \texttt{graphName.part.$k$} and written to the output file.


\subsection{form\_statics}
To collect the per-partition statistics produced by our experiments without recomputing them, we use the script \texttt{form\_statistics}, which processes the files produced by \texttt{calc\_partitions}. For each file, the script reads the contents line by line and parses two types of entries: lines containing \texttt{.part.}, treated as headers marking the beginning of a new partition file's section, and lines starting with \texttt{Pid}, containing the volume of the corresponding partition. The values are grouped by section in a dictionary, and for each group we compute the total outgoing communication volume, the maximum outgoing volume across partitions, the average outgoing volume, and the standard deviation. The results are written to a CSV file, one per input file.

\subsection{graph\_results\_vol}
To evaluate the distribution, we have developed a simple script that plots the partition volumes for a given partition file. The script takes three inputs: the graph file, the number of partitions, and the version identifying the partition file folder. It reads the files output by \texttt{calc\_partitions} and collects all values from the section matching the requested number of partitions. The script is also hardcoded to do the same for the corresponding METIS volumes, and the two sets of volumes are then plotted side by side.

\subsection{stats\_vol}
This script reads the files produced by \texttt{form\_statistics} and plots the results. It takes a graph file name and the version being tested as input, then compares the resulting statistics against the METIS reference values. It produces four plots: the total volume, the maximum partition volume, the percentage difference in maximum partition volume between NVOL and METIS, and the standard deviation of the partition volumes.

\subsection{translate\_to\_metis}
All graphs used in this thesis were in Matrix Market exchange format (.mtx) and therefore needed to be translated into a format METIS could use. The .mtx format is a plain-text format for representing sparse and dense matrices, where each non-zero entry corresponds to an edge in the graph. After the comment lines (prefixed with \%) and a header, each subsequent line lists the two endpoints of an edge: for example, 1 24 denotes an edge between vertex 1 and vertex 24. Our script parses the file and stores each edge as a sorted tuple in a set, which deduplicates entries and discards self-loops. It then builds an adjacency list by iterating over the edge set, appending each vertex to each other's neighbor lists. Since Matrix Market files are 1-indexed while Python lists are 0-indexed, we offset the list indices by one while keeping the vertex ids themselves 1-indexed, as required by the METIS format. Once all edges have been processed, the adjacency list is written as a graph file.

\subsection{model\_estimates} \label{sec:model estimates}
To further evaluate the quality of our partitions, we implemented one of the models presented in~\cite{Xing_of_Heterogeneous}. These models estimate the per-process runtime of a communicating task while accounting for the bandwidth contention caused by competing MPI processes. We specifically use the model for processes with multiple neighbors and non-uniform message sizes, given by formulas 7-9.

Our script takes four arguments: the graph, the version of NVOL, the number of partitions, and the startup latency. The bandwidth array is currently hardcoded, but it could just as easily be exposed as an argument to adapt the model to other setups. Both the startup latency and the bandwidth values were provided by one of the authors of~\cite{Xing_of_Heterogeneous}.

The script begins by computing the incoming communication volume for each partition. From these volumes, together with the bandwidth and latency parameters, we estimate how long each partition takes to complete its communication step. The formulas take the incoming volumes sorted from lowest to highest and evaluate the receive and send times for each partition: formula 7 estimates the total time a partition spends receiving messages as a function of its incoming volume, while formula 9 pinpoints the time to send a message to a specific destination.

Formula 8 then combines these results by taking the maximum of the receive time and the individual send completion times,a partition can only progress as fast as its slowest communication operation, and adds the startup latency multiplied by the number of neighboring partitions. This captures the fact that every message exchange incurs a fixed startup cost regardless of the volume transferred. All three formulas are applied to every partition, and the results are plotted as a bar graph.

\begin{equation}\label{eq:formula8}
	T_i = M^{in}_i \cdot \tau + max\left(t^{\mathrm{recv}}_{i-1}, t^{\mathrm{send}}_{i,0}, t^{\mathrm{send}}_{i,1}, \dots, t^{\mathrm{send}}_{i,M^{\mathrm{out}}_i-1}\right)
\end{equation}