\chapter{Conclusions and Future Work} 
This chapter concludes the thesis, evaluating how well the goals we set in the introduction were reached. We look back at the results and evaluate the work done so far. In the final section we look at future work to be done on this project. 


\section{Conclusions}
In Chapter~\ref{chap:introduction} of this thesis we posed two research questions: 
\begin{itemize}
	\item \textbf{RQ1}: Can a partition-aware objective function produce more balanced communication loads than a global one?
	\item \textbf{RQ2}: Can the communication time of a parallel application be improved in practice by using a partition-aware objective function?
\end{itemize}
With these questions in mind, we have developed and implemented NVOL, a partition-aware objective function that minimizes the maximum per-partition outgoing communication volume. This notion is not new and predates this work; what this thesis contributes is a novel gain calculation to support this objective function, a complete design integrating it into METIS's refinement stage, and an empirical evaluation against METIS's VOL objective. The next section expands on each of these. 

\subsection{Contributions}
The following list contains the contributions provided by this thesis.
\begin{itemize}
	\item  We present a new gain calculation, described in Chapter~\ref{cv_calculation}, that tracks how each partition's volume changes for a given move. This calculation extends the existing total communication volume gain calculation to record the volume changes at the partition level instead of the whole graph. Using the results of this calculation, NVOL is able to evaluate which moves improve the maximum outgoing communication volume. 
	
	\item \textbf{NVOL}, a partition-aware objective function that minimizes the maximum per-partition outgoing communication volume rather than the global total. We present a complete design of this approach in Chapters~\ref{cv_calculation} and~\ref{chapter5}, which we later implement. 
	
	\item An implementation of NVOL within the METIS framework, integrated into the refinement stage with the supporting gain-table data structures it requires. Chapter~\ref{chap:NVOL tooling} describes the structures and functions that were added to METIS in order to implement NVOL. 
		
	\item An empirical study of NVOL against METIS's VOL objective across 20 graphs partitioned into 2 to 2048 parts, on maximum outgoing communication volume, total communication volume and simulated communication time estimates. The graphs were drawn from two sources: the first consists of matrices from heart simulations, and the second is a selection from the SuiteSparse Matrix Collection. NVOL performs better than VOL for most graphs and partition numbers, reaching a peak improvement of $56.70\%$ lower maximum outgoing communication. Throughout the tests, the partitions achieved using NVOL as the objective function have more distributed communication volumes, and in some setups, partitions produced with NVOL have also a lower total communication volume. On the maximum estimated communication-time metric, partitions produced with NVOL are faster in 65\% of our tests.
\end{itemize} 

The next sections evaluate these contributions against the two research questions.
\subsection{RQ1 - Does NVOL improve the volume balance?}
Our results show that using NVOL as the objective function produces partitions with volumes comparable to those produced by METIS using its total communication volume objective function(VOL). NVOL is defined to minimize the maximum outgoing volume ($MV^{\mathrm{max}}$) of any single partition. This allows moves that raise the volume elsewhere as long as the maximum is not increased. NVOL consistently yields partitions with a lower $MV^{\mathrm{max}}$ compared to VOL. The more even distribution is reflected in the lower variance in outgoing volumes, especially in runs with 2 partitions, where NVOL achieves perfectly balanced partitions on almost all graphs.

On near-regular graphs (graphs whose vertices have similar degrees), NVOL consistently improves on $MV^{\mathrm{max}}$. The numerical simulation graphs and the cardiac graphs are particularly notable, with NVOL achieving a lower $MV^{\mathrm{max}}$ than VOL in almost every run. On these graphs, using NVOL also reduces the spread of outgoing volumes across partitions, producing more uniform partition volumes. Notably, on the cardiac graphs and a few others, NVOL even achieves a lower total communication volume than VOL despite not optimizing for that metric.

These results are not universal: NVOL underperforms on certain graph structures. Road network graphs in particular proved challenging in our tests, this was the only graph type where NVOL improves on $MV^{\mathrm{max}}$ only in 55\% of the runs. These graphs have high sparsity, with a few high-degree vertices and many low-degree ones, and this irregular structure prevents NVOL from producing consistent improvements.

The largest graph we tested NVOL on was kmer\_U1a (see Chapter~\ref{chap:results}), a protein k-mer graph with 67.7 million vertices. This graph also has high sparsity, with many small, highly connected components. Using NVOL yields a better $MV^{\mathrm{max}}$ than VOL and a consistently lower standard deviation in the volume values, reaching a reduction in $MV^{\mathrm{max}}$ of up to 46.34\%. This suggests that NVOL performs well when given many options, and raises an interesting question: how well would NVOL perform on even larger graphs? On the negative side, this graph was already too large to partition into 2048 parts with NVOL, making efficiency improvements a priority before scaling further.

This point about NVOL performing better when it has many options is also supported by the results for coPapersDBLP (see Chapter~\ref{chap:results}), a graph with a wide variance in edges per vertex. This structure is common in citation networks, which have a few high-degree hub vertices and many low-degree ones. At high $k$ values, NVOL achieves a peak improvement of $56.70\%$ for $k=512$, with improvements of 40\% or more for all $k \geq 128$ in our tests. NVOL also achieves a lower $MV^{\mathrm{max}}$ than VOL down to $k = 16$. However, for $k < 16$ it underperforms considerably, with the worst result at $-32.09\%$.

We also tested NVOL with imbalance tuning (see Section~\ref{sec:tuning}), which produced more variable results. Allowing greater imbalance allows for more vertex moves, resulting in several additional runs failing to improve on $MV^{\mathrm{max}}$, though NVOL maintained a positive win rate overall. The main effect of relaxed imbalance appears to be a larger magnitude in both wins and losses. The high variance of these results suggests that imbalance tuning should be treated as a per-run adjustment rather than a default setting.

Taken together, these results answer the first question affirmatively: optimizing NVOL produces partitions with more uniform volumes and lower variance compared to METIS's volume objective function. This does push the average partition volume higher, but the increase comes primarily from redistributing the spikes that previously concentrated in a few partitions. Optimizing NVOL improves the maximum outgoing volume by up to $56.70\%$ compared to METIS's total communication volume objective function.

\subsection{RQ2 - Do the partitions produced by NVOL improve communication time?}
The answer to this question is yes, with caveats. NVOL achieves better communication time estimates than VOL in 65\% of our tests, and maintains a 60\% win rate when imbalance tuning is enabled. In winning cases, NVOL improves communication time by an average of 2.61\%, peaking at 6.6\% (without imbalance tuning). 

Some results, such as those for the italy\_osm graph (see Chapter~\ref{chap:results}), also show that message latency can dominate volume in certain settings. Latency is paid per neighboring partition, so a partition with many neighbors but low per-message volume is highly affected by it. As a result, on italy\_osm the partitioning obtained with tuned NVOL had a maximum communication time estimate 27.8\% slower than VOL, even though it achieved a 26.61\% lower $MV^{\mathrm{max}}$. The slowdown came purely from an increase in the number of partition neighbors: with more partitions to message, the per-message latency dominated the overall communication time.

The coPapersDBLP graph highlights another issue: partitions produced with NVOL have a much higher average volume, and this affects communication time negatively. The coPapersDBLP graph shows a great difference in the total volume between partitions produced using NVOL and VOL. This suggests that NVOL might struggle on graphs where its improvement in $MV^{\mathrm{max}}$ comes at the cost of a much higher total communication volume.. 

These results come with important caveats. We were only able to test on an idealized setup with 32 processors, which significantly limited the number of configurations we could explore. To perform more tests, we would need hardware data for larger machines than the one we used. Most of NVOL's greatest improvements appear on partitionings into a large number of parts.  Furthermore, these values are only estimates, providing no definitive confirmation that NVOL's communication times are actually faster on real hardware. 

In summary, NVOL does appear to reduce the communication time in many cases, but these improvements remain simulated. Future work would require additional testing on real machines to validate these findings. Even so, the estimates we do have are promising and suggest that partitions with NVOL as the objective function could lead to meaningful communication time improvements, especially on graphs partitioned into many parts.

\section{Future Work}\label{chap_future_work}
NVOL shows clear promise, achieving improvements over VOL on most of the graphs and configurations we tested, though several limitations and open questions remain. The conclusions above point out some of NVOL's shortcomings. Most notably, its high partitioning time limited the number of tests we were able to perform, and further testing is needed to strengthen the results presented in this thesis. The communication time estimates also point to a promising direction: integrating the communication time model directly into the refinement logic, so that communication time becomes the objective function rather than a proxy through $MV^{\mathrm{max}}$.

\subsection{Improved efficiency} \label{sec:Improved eff}
One of METIS's main strengths is its partitioning speed, and the additional structures and operations introduced by NVOL inevitably affect this, particularly on larger graphs. For small graphs the partitioning time is essentially unchanged, but on larger ones NVOL was observed to be significantly slower than VOL.

The overhead has a straightforward explanation: NVOL adds code to METIS's refinement stage and increases the number of iterations performed there. METIS itself is carefully designed to avoid redundant work, most of the volume operations are performed during graph projection. This way the refinement stage can reuse those results rather than iterating over the graph again. NVOL breaks this pattern by recomputing the gain information for each vertex it visits during refinement, effectively repeating work that METIS already performs at projection time.

Three considerations led us not to integrate NVOL's gain calculation directly into the projection stage. First, NVOL has a larger memory requirement than VOL. As described in Chapter~\ref{cv_calculation}, VOL requires a vector whose length equals the number of partitions a vertex neighbors, whereas NVOL requires a table. For a vertex with a gain vector of length $k$, the corresponding gain table has dimensions $(k+1)\times k$, so a gain vector of length 5 already implies a $6 \times 5$ table per vertex. Second, VOL uses incremental updates between moves to keep the volume gains of affected vertices current; implementing the equivalent logic for NVOL was outside the scope of this thesis. Third, and perhaps most decisive, time constraints during development, combined with the considerable effort required to address the first two points, prevented us from pursuing this integration further. 

This decision was nonetheless preceded by an attempt at the integration. Our strategy allocated the exact memory needed for each gain table and deallocated it once the corresponding vertex left the boundary; gain updates were handled by simply recomputing the gains of all neighbors of the moved vertex. In principle this is much closer to what VOL does and should have improved performance. In practice, however, the algorithm became even slower than the original NVOL implementation, with the frequent allocation and deallocation of memory as the dominant cost. A better solution would be to allocate a single block of memory at projection time, sized from an estimate of how many boundary vertices the finer graph will contain, with additional logic to extend the block when the estimate falls short and to release it once refinement completes. Combined with proper incremental updates of neighboring vertices' gains after each move, this would bring NVOL's partitioning time close to VOL's, at the cost of higher memory use.

\subsection{More tests}
Another clear area of future work is to run more tests on NVOL. In Chapter~\ref{chap:results} we tested only 20 graphs across partition counts ranging from 2 to 2048. NVOL would benefit from being tested on more graphs with different structures, and possibly at higher partition counts.

The imbalance tuning would also benefit from a more detailed study. In Section~\ref{sec:tuning} we tested NVOL only with 4\% imbalance, but it would be informative to increase the imbalance further to observe its effects. Comparing NVOL to VOL when both use the same imbalance setting would also help isolate NVOL's specific contribution.

As mentioned earlier, we tested the communication time model only at $k=32$. Given the data we possess, we can test for all $k \leq 32$, which would help develop more robust conclusions about the communication times of partitions produced with NVOL. It would also be interesting to expand the tests to larger machines, using real hardware values for message latency and network bandwidth. NVOL's best results are at the higher $k$ counts, and it would be interesting to see if that translates over. 

\subsection{Introduce hardware knowledge}
Building on NVOL's the partition-specific implementation, it would be interesting to incorporate some of the logic presented in~\cite{Xing_of_Heterogeneous} to further improve actual communication time. Currently, we partition using communication volume and then model the communication time of each partition based on a set of constants and the incoming volume. With the right adaptations, NVOL could already evaluate this during refinement. Changing NVOL's volume calculations from outgoing to incoming volume is non-trivial, but it has the potential for even larger improvements.

The general idea would be to extend METIS to accept additional parameters describing the message latencies and network bandwidth of the machine running the partitions. Then, during refinement, one could use the equations shown in Section~\ref{sec:model estimates} to evaluate candidate moves. This would increase the chance of producing a partitioning that performs well on the target parallel machine. NVOL's per-partition structure is essential to such an approach.

This could be extended further to optimize for heterogeneous machines. So far we have only considered tests on machines where each component was identical, but in reality modern parallel computers are larger and more complex. Their heterogeneous structure therefore introduces an added level of complexity that can be exploited. In~\cite{Xing_of_Heterogeneous}, the authors also develop models for such setups. With the right structures in place, NVOL could support this logic as well.